---
title: "BaHZING_Model(): Full End-to-End Results Scenario Testing"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    highlight: tango
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{css, echo = FALSE}
pre {
  max-height: 420px;
  overflow-y: auto;
  overflow-x: auto;
}
table.dataTable thead th {
  white-space: normal !important;
  word-break: break-word;
  max-width: 160px;
}
```

```{r libs, message = FALSE}
pkgload::load_all(".", quiet = TRUE)
library(phyloseq)
library(DT)
library(ggplot2)
```

## Overview

Unlike the other two QA docs - `dev/format_bahzing_scenarios.Rmd`
(`Format_BaHZING()`'s data handling) and `dev/bahzing_model_jags_scenarios.Rmd`
(`BaHZING_Model()`'s JAGS *text generation*, verified via `jags.model()`
compile-checks only, no actual sampling) - this one runs the **full
end-to-end pipeline**: format data, fit the model, and produce a real
results table and a real downstream visualization, following the
summarize-and-plot pattern from
[Goodrich-Lab/BaHZING](https://github.com/Goodrich-Lab/BaHZING).

**One fidelity level, uniformly**: every scenario below - including the
default hierarchy - uses the same light MCMC settings (`n.chains = 2,
n.adapt = 60, n.iter.burnin = 2, n.iter.sample = 50`). `n.adapt = 60`
matches `tests/testthat/test-BaHZING_Model.R`'s own already-vetted choice;
`n.iter.sample = 50` (vs. that test file's `2`) avoids a near-degenerate
posterior so the results tables/plots look sensible. This is a demo/showcase
document meant to be re-knitted often while iterating on formatting, so it
stays fast throughout rather than embedding a real, many-minute run.

**Every scenario gets both a results table and a plot**, using whichever
level is the natural analog of "Family" for that hierarchy - not skipped for
shapes without a literal `"Family"`: the default/reduced-depth/middle-only
hierarchies all have `Family` directly; the custom-named hierarchy maps
`Alpha/Bravo/Charlie/Delta/Echo/Foxtrot` position-for-position onto
`Phylum/Class/Order/Family/Genus/Species`, so `Delta` *is* Family's
positional equivalent; the minimum-2-level hierarchy has no mid-level at
all, so it uses `Phylum` (5 taxa - a cleaner plot than Species' 222).

```{r data}
data("iHMP_Reduced")
```

`run_scenario()` builds real formatted data and runs the real
`BaHZING_Model()` with the light settings above - `covar` defaults to
`"consent_age"` so every scenario uses it automatically; `phyloseq_obj`
can be overridden for the custom-named scenario, which needs a phyloseq
object with renamed `tax_table` columns:

```{r run-scenario-helper}
run_scenario <- function(taxa_levels, covar = "consent_age", phyloseq_obj = iHMP_Reduced,
                          x = c("soft_drinks_dietnum", "diet_soft_drinks_dietnum")) {
  formatted_data <- Format_BaHZING(phyloseq_obj, taxa_levels = taxa_levels)
  BaHZING_Model(
    formatted_data = formatted_data,
    x = x,
    covar = covar,
    n.chains = 2, n.adapt = 60, n.iter.burnin = 2, n.iter.sample = 50,
    seed = 20260728, verbose = FALSE
  )
}
```

`summarize_results()` gives a quick sanity check before showing a preview -
total row count, how many *distinct* taxa actually appear (`taxa_full`), and
which level(s) show up in `domain` - cheaper to scan than paging through a
full table, and immediately reveals e.g. one taxon appearing across multiple
component/exposure rows vs. genuinely many distinct taxa:

```{r summarize-results-helper}
summarize_results <- function(result) {
  cat("Total rows:", nrow(result), "\n")
  cat("Unique taxa (taxa_full):", length(unique(result$taxa_full)), "\n")
  cat("Unique domain(s):", paste(unique(result$domain), collapse = ", "), "\n")
  datatable(head(result, 10), options = list(scrollX = TRUE, dom = "t"))
}
```

`plot_domain_results()` is the exact downstream-analysis code from
`Goodrich-Lab/BaHZING`, generalized only enough to reuse across scenarios
(the domain to filter to becomes a parameter, substituted into the
title/axis label) - the filtering/plotting logic itself is unchanged:

```{r plot-domain-helper}
plot_domain_results <- function(bahzing_resout, domain_name) {
  count_estimates <- bahzing_resout[bahzing_resout$component == "Count model coefficients", ]
  domain_res_out <- count_estimates[count_estimates$domain == domain_name, ]
  domain_res_out$p_Dir_sig <- ifelse(domain_res_out$pdir > 0.975, "pDIR Sig", "pDIR Not Sig")

  ggplot(domain_res_out, aes(x = estimate, y = taxa_name, color = p_Dir_sig)) +
    geom_point() +
    geom_errorbar(aes(xmin = bci_lcl, xmax = bci_ucl), width = 0) +
    geom_vline(xintercept = 0, linetype = 2) +
    facet_grid(~exposure) +
    theme_classic() +
    labs(title = paste0(domain_name, "-level count model results"),
         x = "Log fold-change", y = domain_name, color = "Probability of Direction")
}
```

# Scenarios

## 1. Default 6-level hierarchy

```{r scenario-1}
result_1 <- run_scenario(default_taxa_levels)
summarize_results(result_1)
```

The exact downstream-analysis code as originally given (unmodified, using
`bahzing_resout`/`family_res_out` literally, not the generalized helper) -
one section of this document that's precisely the reference pattern, not
just a generalized version of it:

```{r scenario-1-verbatim}
bahzing_resout <- result_1

# Summarize the results
## Lets examine the Count model coefficients
count_estimates <- bahzing_resout[bahzing_resout$component == "Count model coefficients", ]

# Lets look at only the family level results
family_res_out <- count_estimates[count_estimates$domain == "Family",]

# Having a probabilty of direction > 0.975 is equivalent to examining
# whether the 95% BCI is overlapping with 0:
family_res_out$p_Dir_sig <-  ifelse(family_res_out$pdir > 0.975,
                                    "pDIR Sig", "pDIR Not Sig")

# Plot the genus results
ggplot(family_res_out, aes(x = estimate, y = taxa_name, color = p_Dir_sig)) +
  geom_point() +
  geom_errorbar(aes(xmin = bci_lcl, xmax = bci_ucl), width = 0) +
  geom_vline(xintercept = 0, linetype = 2) +
  facet_grid(~exposure) +
  theme_classic() +
  labs(title = "Family-level count model results",
       x = "Log fold-change",
       y = "Family",
       color = "Probability of Direction")
```

## 2. Reduced-depth hierarchy

`c("Phylum", "Family", "Genus", "Species")`.

```{r scenario-2}
result_2 <- run_scenario(c("Phylum", "Family", "Genus", "Species"))
summarize_results(result_2)
plot_domain_results(result_2, "Family")
```

## 3. Middle-only hierarchy

`c("Class", "Order", "Family")`. Note the plot uses `"Order"`, not
`"Family"`: in this hierarchy `Family` is the *narrowest* level - tied 1:1
to each of the 222 original OTUs (like Species in the default hierarchy),
so `domain == "Family"` here matches all 222 individual taxa, not a small
aggregated set. `Order` is the true mid-hierarchy level (has both a parent
to borrow from and a child that borrows from it) and only has 6 distinct
values - the correct analog to how `"Family"` behaves in the scenarios
above.

```{r scenario-3}
result_3 <- run_scenario(c("Class", "Order", "Family"))
summarize_results(result_3)
plot_domain_results(result_3, "Order")
```

## 4. Custom-named hierarchy

`c("Alpha", "Bravo", "Charlie", "Delta", "Echo", "Foxtrot")` - `Delta` is
Family's positional equivalent.

```{r scenario-4}
ps_renamed <- iHMP_Reduced
tt <- tax_table(ps_renamed)
colnames(tt) <- c("Domain", "Alpha", "Bravo", "Charlie", "Delta", "Echo", "Foxtrot")
tax_table(ps_renamed) <- tt

result_4 <- run_scenario(c("Alpha", "Bravo", "Charlie", "Delta", "Echo", "Foxtrot"),
                          phyloseq_obj = ps_renamed)
summarize_results(result_4)
plot_domain_results(result_4, "Delta")
```

## 5. Minimum-2-level hierarchy

`c("Phylum", "Species")` - no mid-level exists, so `Phylum` (5 taxa) is used
instead of `Family`.

```{r scenario-5}
result_5 <- run_scenario(c("Phylum", "Species"))
summarize_results(result_5)
plot_domain_results(result_5, "Phylum")
```

## 6. Default 6-level hierarchy, single exposure, with covariate

Same `default_taxa_levels` hierarchy as Scenario 1, but with only one
exposure (`soft_drinks_dietnum`) instead of two, keeping the default
`consent_age` covariate.

```{r scenario-6}
result_6 <- run_scenario(default_taxa_levels, x = "soft_drinks_dietnum")
summarize_results(result_6)
plot_domain_results(result_6, "Family")
```

## 7. Default 6-level hierarchy, single exposure, without covariate

Same as Scenario 6, but with `covar = NULL` - no covariates in the model.

```{r scenario-7}
result_7 <- run_scenario(default_taxa_levels, covar = NULL, x = "soft_drinks_dietnum")
summarize_results(result_7)
plot_domain_results(result_7, "Family")
```
