---
title: "Format_BaHZING(): Comprehensive Input Scenario Testing"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{css, echo = FALSE}
/* Long taxon-lineage column headers wrap onto multiple lines instead of
   overflowing or forcing the table to scroll unreasonably wide. */
table.dataTable thead th {
  white-space: normal !important;
  word-break: break-word;
  max-width: 160px;
}
```

```{r libs, message = FALSE}
pkgload::load_all(".", quiet = TRUE)
library(phyloseq)
library(DT)
```

## Overview

This is a developer/QA verification document, not end-user tutorial
documentation - it exists to show the *actual, executed* output of
`Format_BaHZING()` across every edge case identified while building and
reviewing the `taxa_levels` generalization refactor, including a few
deliberately-triggered error conditions. It complements (doesn't replace)
the pass/fail assertions in `tests/testthat/test-Format_BaHZING.R` - the
value here is visible, inspectable evidence a human can review directly,
rather than just a green checkmark.

`Format_BaHZING()` alone involves no JAGS/MCMC, so every scenario below is
fast - this whole document builds in well under a few seconds.

```{r data}
data("iHMP_Reduced")
```

A small helper used throughout for the expected-error scenarios, so each
one shows the actual error text without halting the knit:

```{r error-helper}
show_error <- function(expr) {
  result <- tryCatch(expr, error = function(e) e)
  if (inherits(result, "error")) {
    cat("ERROR:", conditionMessage(result), "\n")
  } else {
    cat("No error - unexpected for this scenario.\n")
  }
  invisible(result)
}
```

Two more helpers for previewing actual data content, both using
`DT::datatable()` for a real interactive HTML table (sortable, scrollable)
rather than a static markdown-converted one - long taxon-lineage column
headers wrap across multiple lines via the CSS rule above instead of
overflowing or being abbreviated away.

`preview_table()` shows a handful of metadata columns plus a handful of
taxon-count columns from `Table`, side by side, with their **full** column
names:

```{r preview-table-helper}
preview_table <- function(formatted_data, meta_cols = c("consent_age", "sex", "soft_drinks_dietnum"),
                          n_taxa = 3, n_rows = 5) {
  tbl <- data.frame(formatted_data$Table)
  taxa_cols <- head(formatted_data$taxon_columns, n_taxa)
  preview <- tbl[seq_len(n_rows), c(meta_cols, taxa_cols)]
  datatable(preview, options = list(scrollX = TRUE, dom = "t"), class = "cell-border stripe")
}
```

`preview_matrix()` shows a corner of one of the binary hierarchy incidence
matrices (e.g. `Species.Genus.Matrix`) - rows are the broader level,
columns the narrower level, `1` meaning "this narrower taxon belongs to
this broader taxon" (as established earlier: row/column meaning is fixed
by `hierarchy_matrix_name()`'s own convention, not guessed here):

```{r preview-matrix-helper}
preview_matrix <- function(formatted_data, matrix_name, n_rows = 5, n_cols = 5) {
  m <- formatted_data[[matrix_name]]
  preview <- as.data.frame(m[seq_len(min(n_rows, nrow(m))), seq_len(min(n_cols, ncol(m))), drop = FALSE])
  datatable(preview, options = list(scrollX = TRUE, dom = "t"), class = "cell-border stripe")
}
```

# Hierarchy generalization

## 1. Default 6-level hierarchy (baseline)

No `taxa_levels` argument - falls back to `default_taxa_levels`.

```{r scenario-1}
formatted_default <- Format_BaHZING(iHMP_Reduced)
formatted_default$taxa_levels
names(formatted_default)
length(formatted_default$taxon_columns)
```

```{r scenario-1-preview}
preview_table(formatted_default)
```

Each of the 5 binary hierarchy incidence matrices for the default
hierarchy (rows = broader level, columns = narrower level, `1` = "this
narrower taxon belongs to this broader taxon"):

```{r scenario-1-matrices, results = 'asis'}
matrix_names <- grep("\\.Matrix$", names(formatted_default), value = TRUE)
for (mn in matrix_names) {
  cat("**", mn, "** (", nrow(formatted_default[[mn]]), " x ",
      ncol(formatted_default[[mn]]), ")\n\n", sep = "")
  print(htmltools::tagList(preview_matrix(formatted_default, mn)))
}
```

## 2. Reduced-depth custom hierarchy

Skips `Class` and `Order` entirely.

```{r scenario-2}
reduced_levels <- c("Phylum", "Family", "Genus", "Species")
formatted_reduced <- Format_BaHZING(iHMP_Reduced, taxa_levels = reduced_levels)
formatted_reduced$taxa_levels
names(formatted_reduced)
# Confirm no Class/Order matrices leaked through
any(grepl("Class|Order", names(formatted_reduced)))
```

```{r scenario-2-preview}
preview_table(formatted_reduced)
```

`Family.Phylum.Matrix` - the matrix directly linking `Family` to `Phylum`,
skipping `Class`/`Order` entirely, which is the whole point of this
reduced-depth scenario (contrast with the default hierarchy's separate
`Class.Phylum.Matrix` -> `Order.Class.Matrix` -> `Family.Order.Matrix`
chain):

```{r scenario-2-matrix}
preview_matrix(formatted_reduced, "Family.Phylum.Matrix")
```

## 3. Fully custom-named hierarchy

Renames `tax_table` columns to generic, non-taxonomic labels before running
the full 6-level hierarchy under those names - proves no level name is
hardcoded anywhere, and confirms the prefix-stripping fix (no more stacked
prefixes like `"a__p__..."`). The kingdom-rank column is kept named
`"Domain"` here deliberately (not renamed to `"Kingdom"`) so this scenario
isolates the `taxa_levels`-driven fix alone - see Scenario 11 for the
separate `Kingdom`-named branch case (originally a gap, since fixed).

```{r scenario-3}
ps_renamed <- iHMP_Reduced
tt <- tax_table(ps_renamed)
colnames(tt) <- c("Domain", "Alpha", "Bravo", "Charlie", "Delta", "Echo", "Foxtrot")
tax_table(ps_renamed) <- tt

custom_levels <- c("Alpha", "Bravo", "Charlie", "Delta", "Echo", "Foxtrot")
formatted_custom <- Format_BaHZING(ps_renamed, taxa_levels = custom_levels)
formatted_custom$taxa_levels
names(formatted_custom)
# A clean single prefix per segment throughout, no stacking anywhere:
formatted_custom$taxon_columns[1]
```

```{r scenario-3-preview}
preview_table(formatted_custom)
```

The hierarchy matrices exist here too - just named after the custom levels
(`"Bravo.Alpha.Matrix"` etc.) instead of `"Class.Phylum.Matrix"`, since
`hierarchy_matrix_name()` derives the name from whatever `taxa_levels` was
actually given, not a fixed set of biological rank names:

```{r scenario-3-matrix}
preview_matrix(formatted_custom, "Bravo.Alpha.Matrix")
```

## 4. First-letter collision (expected error)

Two custom levels sharing a first letter - `validate_taxa_levels()` rejects
this rather than silently merging them.

```{r scenario-4}
show_error(Format_BaHZING(iHMP_Reduced, taxa_levels = c("Class", "Clade")))
```

## 4b. Invalid R identifier in `taxa_levels`

Level names get pasted directly into generated JAGS variable names and
`data.frame` column names, so they must be valid R identifiers.
`"16S"` (as in 16S rRNA sequencing - a name a microbiome researcher might
plausibly reach for) starts with a digit, which `make.names()` would
silently rewrite to `"X16S"` - `validate_taxa_levels()` rejects it outright
instead of letting that silent rewrite happen.

```{r scenario-4b}
make.names("16S")
show_error(Format_BaHZING(iHMP_Reduced, taxa_levels = c("16S", "Species")))
```

## 5. Fewer than 2 taxonomic levels

Two genuinely different failure modes that produce similar-sounding but
distinct errors: one checks what you *asked for* (`taxa_levels`), the other
checks what the *data itself* provides, independent of `taxa_levels`.

### 5a. Fewer than 2 elements in the `taxa_levels` argument

```{r scenario-5a}
show_error(Format_BaHZING(iHMP_Reduced, taxa_levels = c("Species")))
```

### 5b. Fewer than 2 columns in the raw input `tax_table`

`taxa_levels` here is left at its normal 6-level default - the input data
itself is what's deficient (only 1 rank column kept, though all 222 taxa/
rows remain intact).

```{r scenario-5b}
PS_one_rank <- iHMP_Reduced
tax_table(PS_one_rank) <- tax_table(tax_table(PS_one_rank)[, 1])
show_error(Format_BaHZING(PS_one_rank))
```

# Data-quality auto-repair

## 6. Narrowest level (Species) column entirely absent

```{r scenario-6}
PS_no_species <- iHMP_Reduced
tax_table(PS_no_species) <- tax_table(PS_no_species)[, colnames(tax_table(PS_no_species)) != "Species"]

formatted_no_species <- Format_BaHZING(PS_no_species)
# Auto-created placeholder names:
head(grep("s__unclassified", formatted_no_species$taxon_columns, value = TRUE), 3)
```

## 7. Mid-hierarchy `NA` values

Three taxa have their `Genus` classification set to `NA` (imagine: sequences
that couldn't be resolved to genus level). These get auto-filled with
`"g__unclassifiedN"` placeholders rather than propagating `NA` into a
downstream column name.

```{r scenario-7}
PS_na_genus <- iHMP_Reduced
tt_na <- tax_table(PS_na_genus)
tt_na[1:3, "Genus"] <- NA
tax_table(PS_na_genus) <- tt_na

formatted_na_genus <- Format_BaHZING(PS_na_genus)
any(is.na(formatted_na_genus$taxon_columns))
grep("g__unclassified", formatted_na_genus$taxon_columns, value = TRUE)
```

## 8. Duplicate names at the narrowest level

Forces two different taxa to share the exact same `Species` classification
(a real occurrence with limited sequencing resolution) - both stay distinct
via a `_N` disambiguating suffix, rather than silently collapsing into one
column.

```{r scenario-8}
PS_dup_species <- iHMP_Reduced
tt_dup <- tax_table(PS_dup_species)
tt_dup[2, "Species"] <- tt_dup[1, "Species"]
tax_table(PS_dup_species) <- tt_dup

formatted_dup <- Format_BaHZING(PS_dup_species)
# Still 222 distinct columns, no silent collapsing:
length(unique(formatted_dup$taxon_columns)) == length(formatted_dup$taxon_columns)
any(grepl("_[0-9]+$", formatted_dup$taxon_columns))

# The actual before/after: row 1 (the original, first occurrence) keeps its
# plain name; row 2 (the forced duplicate) gets the disambiguating "_1" suffix.
cat("Row 1 (original):\n  ", formatted_dup$taxon_columns[1], "\n")
cat("Row 2 (forced duplicate, disambiguated):\n  ", formatted_dup$taxon_columns[2], "\n")
```

One more subtlety visible above, worth being precise about: row 2's
disambiguated name is `"...JAGTTR01sp018223385_1"` (no separator between
`JAGTTR01` and `sp018223385`), while row 1's is `"...JAGTTR01.sp018223385"`
(dot-separated). That's because the duplicate-renaming code
(`str_replace_all(duplicate.names, " ", "")`) strips *all* spaces from the
name it's disambiguating - including the species name's own original
internal space - not just the spaces the renaming step itself introduced.
Non-duplicated names get their internal space turned into a dot later, via
`make.names()`; a duplicated name's internal space is already gone by that
point, so it's simply absent rather than replaced. Still fully unique,
still correct - just a minor cosmetic inconsistency between duplicated and
non-duplicated names, not a correctness issue.

## 9. Kingdom-level `NA` for some taxa (expected error)

Unlike the mid-hierarchy case above, a missing *Kingdom*-level
classification is treated as a hard data-quality error rather than
auto-repaired - the broadest rank is expected to always be resolvable.

```{r scenario-9}
PS_na_kingdom <- iHMP_Reduced
tt_king <- tax_table(PS_na_kingdom)
tt_king[1, "Domain"] <- NA
tax_table(PS_na_kingdom) <- tt_king

show_error(Format_BaHZING(PS_na_kingdom))
```

# Kingdom/Domain handling

## 10. `Domain` column present

This is `iHMP_Reduced`'s actual raw shape - Scenario 1 above already
exercises this path (the SILVA/QIIME2-style `d__` -> `k__` conversion),
included here again for completeness alongside its siblings.

```{r scenario-10}
"Domain" %in% colnames(tax_table(iHMP_Reduced))
# Confirmed already in scenario 1: taxon_columns start with "k__", not "d__"
grepl("^k__", formatted_default$taxon_columns[1])
```

## 11. A literal `"Kingdom"` column instead of `"Domain"`

Real data (like `iHMP_Reduced`) only ever has `Domain`, so this branch
(`if ("Kingdom" %in% colnames(...))`) is otherwise never actually exercised.
**This originally uncovered a real gap**, since fixed: unlike the
`Domain`-specific branch (which explicitly converts `"d__"` -> `"k__"`),
the `Kingdom`-named branch used to only check for an existing `"k__"`
prefix - if the raw values were still SILVA/QIIME2-style (`"d__Bacteria"`),
it prepended `"k__"` without stripping the old prefix, producing a stacked
`"k__d__Bacteria"` instead of clean `"k__Bacteria"`. The same
strip-existing-prefix-first fix already applied to the general
`taxa_levels` loop earlier this session has now been applied here too - see
`tests/testthat/test-Format_BaHZING.R`'s regression test for this exact
scenario.

```{r scenario-11}
PS_literal_kingdom <- iHMP_Reduced
tt_king2 <- tax_table(PS_literal_kingdom)
colnames(tt_king2)[colnames(tt_king2) == "Domain"] <- "Kingdom"
tax_table(PS_literal_kingdom) <- tt_king2

formatted_literal_kingdom <- Format_BaHZING(PS_literal_kingdom)
formatted_literal_kingdom$taxon_columns[1]
# Compare to the clean version from scenario 1:
formatted_default$taxon_columns[1]
```

## 12. Kingdom and Domain both absent entirely

This session's fix: `BaHZING_Model()` used to break silently here (detected
outcome columns via a hardcoded `"k__"` search); `Format_BaHZING()` itself
was already safe.

```{r scenario-12}
PS_no_kingdom <- iHMP_Reduced
tax_table(PS_no_kingdom) <- tax_table(PS_no_kingdom)[, colnames(tax_table(PS_no_kingdom)) != "Domain"]

formatted_no_kingdom <- Format_BaHZING(PS_no_kingdom)
length(formatted_no_kingdom$taxon_columns)
any(grepl("k__", formatted_no_kingdom$taxon_columns))
formatted_no_kingdom$taxon_columns[1]
```

# Structural edge cases

## 13. `taxa_are_rows = TRUE` input orientation

`iHMP_Reduced` natively has `taxa_are_rows() == FALSE` (samples as rows).
This confirms the very first check in `Format_BaHZING()` (transpose to a
consistent orientation before doing anything else) actually works -
previously unexercised by any test this session.

```{r scenario-13}
taxa_are_rows(iHMP_Reduced)
PS_transposed <- t(iHMP_Reduced)
taxa_are_rows(PS_transposed)

formatted_transposed <- Format_BaHZING(PS_transposed)
identical(formatted_default$Table, formatted_transposed$Table)
identical(formatted_default$Species.Genus.Matrix, formatted_transposed$Species.Genus.Matrix)
identical(formatted_default$taxon_columns, formatted_transposed$taxon_columns)
```

## 14. Already-prefixed vs. unprefixed raw taxonomy values

`iHMP_Reduced`'s raw values already carry `p__`/`c__`/etc. (scenario 1
already demonstrates this, implicitly). Here, those prefixes are stripped
out entirely before running `Format_BaHZING()`, confirming both converge to
the identical, correctly-prefixed result either way.

```{r scenario-14}
PS_stripped <- iHMP_Reduced
tt_strip <- tax_table(PS_stripped)
for (col in colnames(tt_strip)) {
  tt_strip[, col] <- sub("^[a-zA-Z]__", "", tt_strip[, col])
}
tax_table(PS_stripped) <- tt_strip
# Raw values no longer carry any prefix:
as.character(tax_table(PS_stripped)[1, "Genus"])

formatted_stripped <- Format_BaHZING(PS_stripped)
identical(formatted_stripped$taxon_columns, formatted_default$taxon_columns)
```

# Output contract check

## 15. `taxon_columns` matches `Table`'s actual column names

The consistency this session's `make.names()` fix established - confirmed
across a few of the scenarios above, not just the default case.

```{r scenario-15}
check_consistency <- function(formatted_data, label) {
  ok <- all(formatted_data$taxon_columns %in% colnames(data.frame(formatted_data$Table)))
  cat(label, ": ", ok, "\n", sep = "")
}

check_consistency(formatted_default, "Default (scenario 1)")
check_consistency(formatted_reduced, "Reduced depth (scenario 2)")
check_consistency(formatted_custom, "Custom names (scenario 3)")
check_consistency(formatted_no_species, "Species absent (scenario 6)")
check_consistency(formatted_no_kingdom, "Kingdom/Domain absent (scenario 12)")
check_consistency(formatted_transposed, "Transposed input (scenario 13)")
```

# Summary

All scenarios above ran to completion with the expected result - either
successful, correctly-repaired output, or a clear error for genuinely
invalid input. Scenario 11 (a literal `"Kingdom"` column carrying
un-converted `"d__"`-style values) originally exposed a real gap - a
stacked prefix instead of a clean one, since real data only ever exercises
the `Domain`-named branch, not this one - which has since been fixed in
`Format_BaHZING()` and is now covered by a dedicated regression test.
