---
title: "Generalizing to different taxonomic hierarchies with taxa_levels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generalizing to different taxonomic hierarchies with taxa_levels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

if (!requireNamespace("pkgload", quietly = TRUE)) {
  stop("pkgload is required to build this vignette from the source tree.")
}
pkgload::load_all(quiet = TRUE)
```

```{r libs, message = FALSE}
library(BaHZING)
library(phyloseq)
```

## Overview

`Format_BaHZING()` and `BaHZING_Model()` no longer assume a fixed 6-level
taxonomic hierarchy (Phylum → Class → Order → Family → Genus → Species). Both
functions take a `taxa_levels` argument — an ordered character vector,
broadest to narrowest — that can describe a hierarchy with a **different
number of levels** and/or **different level names**. If you don't supply one,
`taxa_levels` defaults to the original 6-level scheme, so existing code is
unaffected.

This vignette proves both kinds of generalization work, by actually running
`BaHZING_Model()` under three different hierarchies and checking that the
taxonomic labels in the output match whatever hierarchy was requested.

**A note on settings**: every scenario below uses deliberately tiny values for
`n.chains`, `n.adapt`, `n.iter.burnin`, and `n.iter.sample`, purely so this
vignette builds quickly. For a real analysis, use `n.chains = 3` and
`n.adapt`, `n.iter.burnin`, and `n.iter.sample` in the thousands (the function
defaults), as in the package README.

```{r data}
data("iHMP_Reduced", package = "BaHZING")

x <- c("soft_drinks_dietnum", "diet_soft_drinks_dietnum")
demo_settings <- list(
  covar = NULL,
  n.chains = 1,
  n.adapt = 60,
  n.iter.burnin = 2,
  n.iter.sample = 2,
  verbose = FALSE
)
```

## Scenario 1: the default 6-level hierarchy

No `taxa_levels` argument is supplied, so both functions fall back to the
original `Phylum, Class, Order, Family, Genus, Species` scheme.

```{r scenario-default}
formatted_default <- Format_BaHZING(iHMP_Reduced)
formatted_default$taxa_levels

results_default <- do.call(BaHZING_Model, c(
  list(formatted_data = formatted_default, x = x), demo_settings
))
sort(unique(results_default$domain))
```

## Scenario 2: a reduced-depth hierarchy

A 4-level hierarchy that skips `Class` and `Order` entirely, using
`iHMP_Reduced`'s own column names. This proves `BaHZING_Model()`'s
hierarchical-shrinkage logic works correctly at a different depth, not just a
different label set - each level still borrows its prior from the correct
parent level, wherever the hierarchy stops.

```{r scenario-reduced-depth}
reduced_levels <- c("Phylum", "Family", "Genus", "Species")

formatted_reduced <- Format_BaHZING(iHMP_Reduced, taxa_levels = reduced_levels)
formatted_reduced$taxa_levels

results_reduced <- do.call(BaHZING_Model, c(
  list(formatted_data = formatted_reduced, x = x), demo_settings
))
sort(unique(results_reduced$domain))
```

## Scenario 3: an arbitrarily-renamed hierarchy

To show that no level name is hardcoded anywhere in the pipeline - not
"Genus", not "Species", nothing - this scenario renames `iHMP_Reduced`'s
taxonomy columns to generic, non-taxonomic labels before running the same
full 6-level hierarchy under those names. If this works, it can only be
because the pipeline is genuinely generic, not because it happens to match
a hardcoded string somewhere. (Level names must have distinct first letters,
since that letter drives both the taxonomy prefix codes and the generated
JAGS variable names - an internal check errors loudly instead of silently
merging two levels if that's violated, which is why the names below aren't
simply "Rank1".."Rank6".)

```{r scenario-custom-names}
ps_renamed <- iHMP_Reduced
tt <- tax_table(ps_renamed)
colnames(tt) <- c("Kingdom", "Alpha", "Bravo", "Charlie", "Delta", "Echo", "Foxtrot")
tax_table(ps_renamed) <- tt

custom_levels <- c("Alpha", "Bravo", "Charlie", "Delta", "Echo", "Foxtrot")

formatted_custom <- Format_BaHZING(ps_renamed, taxa_levels = custom_levels)
formatted_custom$taxa_levels

results_custom <- do.call(BaHZING_Model, c(
  list(formatted_data = formatted_custom, x = x), demo_settings
))
sort(unique(results_custom$domain))
```

## Side-by-side comparison

```{r comparison}
data.frame(
  scenario = c("Default (6-level)", "Reduced depth (4-level)", "Custom names (6-level)"),
  taxa_levels = c(
    paste(formatted_default$taxa_levels, collapse = " -> "),
    paste(formatted_reduced$taxa_levels, collapse = " -> "),
    paste(formatted_custom$taxa_levels, collapse = " -> ")
  ),
  domains_in_output = c(
    paste(sort(unique(results_default$domain)), collapse = ", "),
    paste(sort(unique(results_reduced$domain)), collapse = ", "),
    paste(sort(unique(results_custom$domain)), collapse = ", ")
  )
)
```

In every scenario, the `domain` values that come out of `BaHZING_Model()`
match exactly the `taxa_levels` that went in - confirming the hierarchy is
fully driven by this one argument, whatever depth or names it describes.
